Normalization by Evaluation for Call-By-Push-Value and Polarized Lambda Calculus

### 2019-05-10 ppdp19/ppdp19.tex
Typos.

\author{Christian Sattler}
\affiliation{%
  \institution{Nottingham University}
  \streetaddress{}
  \city{Nottingham}
  \country{UK}
}
%\email{christian.sattler@gmail.com}

@@

Since positive types $P$ have a monadic interpretation, there is a
natural transformation $\C\,\den P \todot \den P$, which is simply
$\tjoin[\C]$.  It
can be extended to a
natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
type $A$, meaning we can \emph{run} the monad,
pushing its effects into $\den A$.%
\footnote{In categorical terminology, the existence of
$\trun^A$ means that
all semantic types $\den{A}$ fulfill the \emph{weak sheaf condition},
aka \emph{weak pasting}.}
% All semantic types fulfill the weak sheaf condition aka weak pasting,
% meaning there is
% a natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
% type $A$.  In other words, we can \emph{run} the monad, pushing its
% effects into $\den A$.
We proceed by induction on $A$.
At negative types we can recurse pointwise at a smaller type,
exploiting that values of negative types are essentially
(finite or infinite) tuples.
\[
\begin{array}{l@{~~}l@{~~}c@{~~}l}%{l@{~}l@{~~}c@{~~}l}
  \multicolumn 4 l {\trun^A ~:~ \C \den A \todot \den A} \\
  \trun^P & c & = & \tjoin[\C]\;c \\
  \trun^{\tyone} & c & = & \ttempty
    \\
  \trun^{A \times B} & c & = &
      (\trun^A\,(\tmap[\C]\;\pi_1\;c),\
       \trun^B\,(\tmap[\C]\;\pi_2\;c))
  \\
  \trun^{A \To B} & c  & = & \lambda\,_\Delta \,\tau \; a. \ \trun^B\,
      (\stmap[\C]\;(\lambda\,_\Phi\;\tau'\,f.\;f\;\tid\;(\tren\;\tau'\;a))
  \\ &&& \qquad \qquad (\tren\;\tau\;c))
  \\
  % \trun^{o}      & & = & \tjoin[\C]
  % \\
  % \trun^{\tyzero} & & = & \tjoin[\C]
  % \\
  % \trun^{A + B}   & & = & \tjoin[\C]
  % \\
\end{array}
\]
For the case of function types $A \To B$, we require the monad $\C$ to be
\emph{strong}, which amounts to having
$\stmap[\C]_\Gamma\, \ell : \C\A\,\Gamma \to \C\,\B\,\Gamma$ already for
a ``local'' presheaf morphism $\ell : (\A \hatto \B)\Gamma$.
The typings are
$c : \C \den{A \To B} \Gamma$ and
$\tau : \Gamma \sublist \Delta$ and
$a : \den A \Delta$, and now we want to apply every function
$f : \den{A \To B}$ in the cover $c$ to argument $a$.  Clearly,
$\tmap[\C]$ is not applicable since it would expect a
global presheaf morphism
$\den{A \to B} \todot \den B$, \ie, something that works in \emph{any}
context.  However, applying to $a : \den A \Delta$ can only work in
context $\Delta$ or any extension $\tau' : \Delta \sublist \Phi$,
since we can transport $a$ to such a $\Phi$ via
$a' := \tren\; \tau'\; a : \den A \Phi$, but not to a context unrelated to
$\Delta$.  We obtain our input to $\trun^B$ of type $\C \den B \Gamma$
as an instance of $\stmap[\C]_\Gamma$ applied to the local presheaf morphism
$(\lambda\,_\Phi \; \tau'\, f. \  f\;\tid\;a') :
\Delta \sublist \Phi \to  \den{A \To B}\Phi \to \den B \Phi$
% (\tau' : \Delta \sublist \Phi) (f : \den{A \To B}\Phi) \to \den B \Phi$
and the transported cover $\tren\;\tau\;c : \C\den{A \To B}\Delta$.

@@

CBPV calls positive types $P$ \emph{value types} $A$ and negative
types $N$
\emph{computation types} $\underline B$,
yet we shall stick to our terminology
which is common in publications on focalization.
However, we shall use $\tThunk$ for switch $\downarrow$ and $\tComp$ for
switch $\uparrow$, to avoid confusion with our notation for reflection
and reification.
\[
\begin{array}{l@{~~}l@{~~}l@{~~}r@{~~}l}%@{\qquad}l}
  \PTy & \ni & P & ::= & \patom
    \mid \tyone \mid P_1 \times P_2
    \mid \tyzero \mid P_1 + P_2
    \mid \Thunk N
%  & \mbox{positive type}
\\
  \NTy & \ni & N & ::= & \natom
    \mid \top \mid N_1 \ntimes N_2
    \mid P \To N
    \mid \Comp P
%  & \mbox{negative type}
\\
\end{array}
\]
CBPV uses $U$ for $\tThunk$ and $F$ for $\tComp$, however, we find
these names uninspiring unless you have good knowledge of the intended
model.  Further, CBPV \cite{levy:hosc06}
employs labeled sums $\Sigma_I (P_i)_{i : I}$
and labeled records $\Pi_I (P_i)_{i : I}$ for up to countably infinite
label sets $I$ while we only have finite sums $(\tyzero,+)$ and
records $(\top,\ntimes)$.  However, this difference is not essential,
our treatment extends directly to the infinite case, since we are
working in type theory which allows infinitely branching inductive
types.   As a last difference, CBPV does not consider base types; in
anticipation of the next section, we add them as both positive atoms
($\patom$) and negative atoms ($\natom$).

@@

Reification $\down P : \den P \todot \tVNf\,P$ at positive types $P$ produces
a normal value, and $\down N : \den N \todot \tNf\,N$ at negative
types $N$ a
normal term.  During reification of function types $P \To N$ in
context $\Gamma$ we need to embed a fresh variable
$x : P \in (\cext \Gamma P)$ into $\den P$, breaking down $P$ to
positive atoms $\patom$ and negative remainders $\Thunk N$.
However, in $\den P$ we do not have case analysis available, thus,
positive reflection $\upG P : P \in \Gamma \to \Cover {\den P} \Gamma$
needs to run in the monad.
Luckily, we can unwind the monad using $\trun^N$ before we recursively
reify at type $N$.
% %% TODO
% Luckily, as $\down N$ produces a normal
% form, monadic intermediate computations are permitted under a final
% $\trunNf$.
Negative reflection $\up N : \tNe\,N \todot \den N$ is as
before generalized from variables to neutrals, to handle the breaking
down of $N$ via eliminations.
%
In the following definition of reflection we use the abbreviation
\fbox{\(
  \tfresh^P_\Gamma
    = \up[\cext \Gamma P] P \vv0
    : \Cover{\den P}{(\cext \Gamma P)}
\)}.
\[
\def\arraystretch{1.3}
\begin{array}{l@{\,}lcl}%@{~~~}c@{~~~}l}
  \multicolumn 2 l {\upG P} & : & P \in \Gamma \to \Cov\,\den P \, \Gamma\\
  \upG {\patom}         & x & = & x \\
  \upG {\tyone}         & x & = & \treturn\;\ttempty \\
  \upG {P_1 \times P_2} & x & = & \tsplit\;x\,
    \left(
     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
     \star
     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
    \right) \\
    % (\tzip
    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
    % ) \\
  \upG{\tyzero}         & x & = & \tabort\;x \\
  \upG{P_1 + P_2}       & x & = & \tcase\;x
    \; (\tmap\;\iota_1\;\tfresh^{P_1}_\Gamma)
    \; (\tmap\;\iota_2\;\tfresh^{P_2}_\Gamma)
    \\
  \upG{\Thunk N} & x & = & \treturn\,(\upG N (\tforce\,x))
\\[2ex]
\end{array}
\]
\[
\def\arraystretch{1.3}
\begin{array}{l@{\,}lcl}
    \multicolumn 2 l {\down P} & : & \den P \todot \tVNf\,P \\
  \down {\patom} & & = & \tvar \\
  \downG {\tyone} & () & = & \punit \\
  \downG {P_1 \times P_2} & (a_1,a_2) & = & \ppair
    \,(\downG {P_1} a_1)
    \,(\downG {P_2} a_2)
    \\
  \down {\tyzero} & & = & \tmagic \\
  \down {P_1 + P_2} & & = &
    [  \tinj_1 \comp \down{P_1}
    ,\ \tinj_2 \comp \down{P_2}
    ] \\
  \down {\Thunk N} & & = & \tthunk \circ \down N \\
\end{array}
\]
Reflection at positive pairs uses monoidal functoriality
$\C\,\A_1 \todot \C\,\A_2 \todot \C\,(\A_1 \hattimes \A_2)$
called $\star$ by McBride and Paterson \cite[Section 7]{mcbridePaterson:applicative}
which for monads $\C$ can be defined by
$c_1 \star c_2 =
   \tjoin \; (\tmap\;(\lambda\,a_1.\
   \tmap\;(\lambda\,a_2.\ (a_1,a_2))
    \;c_2)
    \;c_1)$.
% $c_1 \star c_2 =
%    \tjoin \; (\tmap\;(\lambda\,a_1.\
%    \tjoin \; (\tmap\;(\lambda\,a_2.\ \treturn\,(a_1,a_2))
%     \;c_2))
%     \;c_1)$.
% $c \star d = \tjoin \; (\tmap\;(\lambda\,b.\ (c,\, \treturn\,b))\;d$.
% $\tzip\;c = \tjoin \comp \tmap\,(\lambda\,a.\ (c,\, \treturn\,a)$.

@@

Due to the eager splitting of positive hypotheses, reflection at type $P$
now lives in the graded monad $\Add P$ rather than $\Cov$.  Further,
as pattern matching may produce $n \geq 0$ cases, reflection cannot simply
produce a single positive semantic value; instead, one such value is needed
for every branch.  We implement
\fbox{$\reflect P : (\den P \hatto \J) \todot \Add P \J$}
as a higher-order function expecting a continuation $k$ which is
invoked for each generated branch with the semantic value of type $P$
constructed for this branch.
\[
\renewcommand{\left}{}
\renewcommand{\right}{}
\begin{array}{l@{~}lcl}
  \reflect {\patom} & k & = &
    \phyp \left( k\;\twk^{\patom}\,(\pvar\,\tzero) \right) \\
  \reflect {\Thunk N} & k & = &
    \nhyp \left( k\;\twk^N \left( \up N (\nvar\,\tzero) \right) \right)\\
  \reflect {\tyzero} & k & = & \nbranch \\
  \reflect {P_1 + P_2} & k & = & \bbranch
    \,\left( \treflect^{P_1}\,(\lambda\,\tau. \ k\,\tau \comp \iota_1) \right)
\\ &&& \qquad \quad \ \,
    \,\left( \treflect^{P_2}\,(\lambda\,\tau. \ k\,\tau \comp \iota_2) \right)
    \\
  \reflect{\tyone} & k & = & \nsplit\;(k\;\tid\;\ttempty) \\
  \reflect{P_1 \times P_2} & k & = & \bsplit
    \,\left(
      \treflect^{P_1} \left( \lambda\,\tau_1\,a_1. \
\\ &&& \qquad \ \
      \treflect^{P_2} \left( \lambda\,\tau_2\,a_2. \
\\ &&& \qquad \ \
        k\, (\tau_1 \rcomp \tau_2)\, (\tren\,\tau_2\,a_1, \ a_2)
        \right)
      \right)
    \right)
\end{array}
\]
Reflecting at a positive atomic type $\patom$ is the regular ending of a
reflection pass: we call continuation $k$ with a fresh variable
$\pvar\,\tzero$ of type $\patom$, making space for the variable
using $\twk^{\patom}$.
In case we end at type $\Thunk N$, we add a new variable
$\nvar\,\tzero$ of type $N$ and pass it to $k$, after full
$\eta$-expansion via $\up N$.  Two more endings are possible:
At type
$\tyzero$, we have reached an absurd case, meaning that no
continuation is necessary since we can conclude with
\emph{ex falso quodlibet}.
At type $\tyone$, there is no need to add a new variable, as values
of type $\tyone$ contain no information.  We simply pass the unit
value $\ttempty$ to $k$ in this case.
Reflecting at $P_1 + P_2$ generates two branches, which may result in
several uses of the continuation $k$.
In the first branch, we recursively reflect at $P_1$.  Its
continuation will receive a semantic value in $\den{P_1}$ which we
inject via $\iota_1$ into $\den{P_1 + P_2}$ to pass it to $k$.  The
second branch proceeds analogously.
Finally reflecting at $P_1 \times P_2$ means we first have to analyze
$P_1$, and in each of the generated branches we continue to analyze
$P_2$.  Thus $\reflect{P_2}$ is passed as a continuation to
$\reflect{P_1}$.  Each reflection phase gives us a semantic value
$a_i$ of type $P_i$, which we combine to a tuple before passing it to
$k$.  Note also that the context extension $\tau_1$ created in the
first phase needs to be composed with the context extension $\tau_2$
of the second phase to transport $k$ into the final context.  Further,
the value $a_1$ was constructed relative to the target of
$\tau_1$ and still needs
to be transported with $\tau_2$ before paired up with $a_2$.

@@

The terms \fbox{$\Tm N \Gamma$} of the focused lambda calculus are the
ones of CBPV minus the positive eliminations ($\tsplit$, $\tcase$,
$\tabort$), the added negative variable rule ($\nvar$) and the
necessary changes to the binders $\tabs$ and $\tbind$.
\begin{gather*}
  \nru{\nvar}{N \in \Gamma}{\Tm N \Gamma}
\qquad
  \nru{\tabs}
      {\AddHypP P {\tTm\, N} \Gamma}
      {\TmP {P \To N} \Gamma}
\\[2ex]
  \nru{\tbind}
      {\TmP {\Comp P} \Gamma \qquad \AddHypP P {\tTm\,N} \Gamma}
      {\Tm N \Gamma}
\end{gather*}
Term interpretation
\fbox{$\denty\_ : \Tm N \Gamma \to \den \Gamma \todot \den N$}
shall be as for CBPV except that we need to exchange the
interpretation function for binders
%\fbox{$
\[
\fden\_ : \Tm N {(\cext
    \Gamma P)} \to \den \Gamma \todot \den{P \To N}
.\]
%$}.
Since a binder
for $P$ performs a maximal splitting on $P$ and takes the form of a
function defined by a case (and split) tree,
applying it to a value $v$ of type $P$ amounts to a complete
matching of $v$ against the case tree, and bind the remaining
atomic and negative crumbs.  This matching can be defined for a
generic evaluation function of type
$\Evvv \J \Delta \Gamma = \den \Gamma \Delta \to \J\,\Delta$.
\[
\begin{array}{l@{~}l@{~}l@{~}lcl}
  \multicolumn 6 {l} {\tmatch ~~:~~ \den P \Delta \to
    \AddHypP P {\Evv \J \Delta} \todot \Evv \J \Delta }\\
  \tmatch & x & (\phyp\,e) & \gamma & = & e\,(\gamma,\; \treturn\,x) \\
  \tmatch & b & (\nhyp\,e) & \gamma & = & e\,(\gamma,\; b) \\
  \tmatch & a & \nbranch & \gamma & = & \tmagic\;a \\
  \tmatch & (\iota_1\;a_1) & (\bbranch\;e_1\;e_2) & \gamma & = &
    \tmatch\;a_1\;e_1\;\gamma \\
  \tmatch & (\iota_2\;a_2) & (\bbranch\;e_1\;e_2) & \gamma & = &
    \tmatch\;a_2\;e_2\;\gamma \\
  \tmatch & \ttempty & (\nsplit\,e) & \gamma & = & e\,\gamma \\
  \tmatch & (a_1,a_2) & (\bsplit\,e) & \gamma & = &
\\ \multicolumn 6 {@{\qquad}l}{
    \tmatch\;a_1\;(\tmap[\Add P]\,(\tmatch\;a_2)\;e)\;\gamma
}
\\
\end{array}
\]
With instantiations $\J = \den N$ and
$\denty\_ : \Tm N \todot \Evv {\den N} \Delta$,
the interpretation $\fden t$
of binder $t : \AddHypP P {\tTm\,N} \Gamma$
is defined as follows:
\begin{multline*}
  \fdent{t}
    {(\gamma : \den \Gamma \Delta)}
    \, (\tau : \Delta \sublist \Phi)
    \, (a : \den P \Phi)
    ~~=~~
\\ \qquad
    \tmatch
      \; a
      \; (\tmap[\Add P]\;\denty\_\;t)
      \; (\tren\,\tau\,\gamma)
\end{multline*}
This completes the definition of the normalization function
$\tnorm\, (t : \Tm N \Gamma)
  = \down N \dent t {\idenv\Gamma}$.

>>>>

\author{Christian Sattler}
\affiliation{%
  \institution{School for Computer Science, University of Nottingham}
  \streetaddress{}
  \city{Nottingham}
  \country{United Kingdom}
}
%\email{christian.sattler@gmail.com}

@@

Since positive types $P$ have a monadic interpretation, there is a
natural transformation $\C\,\den P \todot \den P$, which is simply
$\tjoin[\C]$.  It
can be extended to a
natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
type $A$, meaning we can \emph{run} the monad,
pushing its effects into $\den A$.%
\footnote{In categorical terminology, the existence of
$\trun^A$ means that
all semantic types $\den{A}$ fulfill the \emph{weak sheaf condition},
aka \emph{weak pasting}.}
% All semantic types fulfill the weak sheaf condition aka weak pasting,
% meaning there is
% a natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
% type $A$.  In other words, we can \emph{run} the monad, pushing its
% effects into $\den A$.
We proceed by induction on $A$.
At negative types we can recurse pointwise at a smaller type,
exploiting that values of negative types are essentially
(finite or infinite) tuples.
\[
\begin{array}{l@{~~}l@{~~}c@{~~}l}%{l@{~}l@{~~}c@{~~}l}
  \multicolumn 4 l {\trun^A ~:~ \C \den A \todot \den A} \\
  \trun^P & c & = & \tjoin[\C]\;c \\
  \trun^{\tyone} & c & = & \ttempty
    \\
  \trun^{A \times B} & c & = &
      (\trun^A\,(\tmap[\C]\;\pi_1\;c),\
       \trun^B\,(\tmap[\C]\;\pi_2\;c))
  \\
  \trun^{A \To B} & c  & = & \lambda\,_\Delta \,\tau \; a. \ \trun^B\,
      (\stmap[\C]\;(\lambda\,_\Phi\;\tau'\,f.\;f\;\tid\;(\tren\;\tau'\;a))
  \\ &&& \qquad \qquad (\tren\;\tau\;c))
  \\
  % \trun^{o}      & & = & \tjoin[\C]
  % \\
  % \trun^{\tyzero} & & = & \tjoin[\C]
  % \\
  % \trun^{A + B}   & & = & \tjoin[\C]
  % \\
\end{array}
\]
For the case of function types $A \To B$, we require the monad $\C$ to be
\emph{strong}, which amounts to having
$\stmap[\C]_\Gamma\, \ell : \C\A\,\Gamma \to \C\,\B\,\Gamma$ already for
a ``local'' presheaf morphism $\ell : (\A \hatto \B)\Gamma$.
The typings are
$c : \C \den{A \To B} \Gamma$ and
$\tau : \Gamma \sublist \Delta$ and
$a : \den A \Delta$, and now we want to apply every function
$f : \den{A \To B}$ % TODO: this doesn't really typecheck...
in the cover $c$ to argument $a$.  Clearly,
$\tmap[\C]$ is not applicable since it would expect a
global presheaf morphism
$\den{A \to B} \todot \den B$, \ie, something that works in \emph{any}
context.  However, applying to $a : \den A \Delta$ can only work in
context $\Delta$ or any extension $\tau' : \Delta \sublist \Phi$,
since we can transport $a$ to such a $\Phi$ via
$a' := \tren\; \tau'\; a : \den A \Phi$, but not to a context unrelated to
$\Delta$.  We obtain our input to $\trun^B$ of type $\C \den B \Gamma$
as an instance of $\stmap[\C]_\Gamma$ applied to the local presheaf morphism
$(\lambda\,_\Phi \; \tau'\, f. \  f\;\tid\;a') :
\Delta \sublist \Phi \to  \den{A \To B}\Phi \to \den B \Phi$
% (\tau' : \Delta \sublist \Phi) (f : \den{A \To B}\Phi) \to \den B \Phi$
and the transported cover $\tren\;\tau\;c : \C\den{A \To B}\Delta$.

@@

CBPV calls positive types $P$ \emph{value types} $A$ and negative
types $N$
\emph{computation types} $\underline B$,
yet we shall stick to our terminology
which is common in publications on focalization.
However, we shall use $\tThunk$ for switch $\downarrow$ and $\tComp$ for
switch $\uparrow$, to avoid confusion with our notation for reflection
and reification.
\[
\begin{array}{l@{~~}l@{~~}l@{~~}r@{~~}l}%@{\qquad}l}
  \PTy & \ni & P & ::= & \patom
    \mid \tyone \mid P_1 \times P_2
    \mid \tyzero \mid P_1 + P_2
    \mid \Thunk N
%  & \mbox{positive type}
\\
  \NTy & \ni & N & ::= & \natom
    \mid \top \mid N_1 \ntimes N_2
    \mid P \To N
    \mid \Comp P
%  & \mbox{negative type}
\\
\end{array}
\]
CBPV uses $U$ for $\tThunk$ and $F$ for $\tComp$, however, we find
these names uninspiring unless you have good knowledge of the intended
model.  Further, CBPV \cite{levy:hosc06}
employs labeled sums $\Sigma_I (P_i)_{i : I}$
and labeled records $\Pi_I (P_i)_{i : I}$ for up to countably infinite
label sets $I$ while we only have finite sums $(\tyzero,+)$ and
records $(\top,\ntimes)$.  However, this difference is not essential,
our treatment extends directly to the infinite case, since we are
working in type theory, which allows infinitely branching inductive
types.   As a last difference, CBPV does not consider base types; in
anticipation of the next section, we add them as both positive atoms
($\patom$) and negative atoms ($\natom$).

@@

Reification $\down P : \den P \todot \tVNf\,P$ at positive types $P$ produces
a normal value, and $\down N : \den N \todot \tNf\,N$ at negative
types $N$ a
normal term.  During reification of function types $P \To N$ in
context $\Gamma$ we need to embed a fresh variable
$x : P \in (\cext \Gamma P)$ into $\den P$, breaking down $P$ to
positive atoms $\patom$ and negative remainders $\Thunk N$.
However, in $\den P$ we do not have case analysis available, thus,
positive reflection $\upG P : P \in \Gamma \to \Cover {\den P} \Gamma$
needs to run in the monad.
Luckily, we can unwind the monad using $\trun^N$ before we recursively
reify at type $N$.
% %% TODO
% Luckily, as $\down N$ produces a normal
% form, monadic intermediate computations are permitted under a final
% $\trunNf$.
Negative reflection $\up N : \tNe\,N \todot \den N$ is as
before generalized from variables to neutrals, to handle the breaking
down of $N$ via eliminations.
%
In the following definition of reflection we use the abbreviation
\fbox{\(
  \tfresh^P_\Gamma
    = \up[\cext \Gamma P] P \vv0
    : \Cover{\den P}{(\cext \Gamma P)}
\)}.
\[
\def\arraystretch{1.3}
\begin{array}{l@{\,}lcl}%@{~~~}c@{~~~}l}
  \multicolumn 2 l {\upG P} & : & P \in \Gamma \to \Cov\,\den P \, \Gamma\\
  \upG {\patom}         & x & = & x \\
  \upG {\tyone}         & x & = & \treturn\;\ttempty \\
  \upG {P_1 \times P_2} & x & = & \tsplit\;x\,
    \left(
     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
     \star
     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
    \right) \\
    % (\tzip
    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
    % ) \\
  \upG{\tyzero}         & x & = & \tabort\;x \\
  \upG{P_1 + P_2}       & x & = & \tcase\;x
    \; (\tmap\;\iota_1\;\tfresh^{P_1}_\Gamma)
    \; (\tmap\;\iota_2\;\tfresh^{P_2}_\Gamma)
    \\
  \upG{\Thunk N} & x & = & \treturn\,(\upG N (\tforce\,x))
\\[2ex]
\end{array}
\]
\[
\def\arraystretch{1.3}
\begin{array}{l@{\,}lcl}
    \multicolumn 2 l {\down P} & : & \den P \todot \tVNf\,P \\
  \down {\patom} & & = & \tvar \\
  \downG {\tyone} & () & = & \punit \\
  \downG {P_1 \times P_2} & (a_1,a_2) & = & \ppair
    \,(\downG {P_1} a_1)
    \,(\downG {P_2} a_2)
    \\
  \down {\tyzero} & & = & \tmagic \\
  \down {P_1 + P_2} & & = &
    [  \tinj_1 \comp \down{P_1}
    ,\ \tinj_2 \comp \down{P_2}
    ] \\
  \down {\Thunk N} & & = & \tthunk \circ \down N \\
\end{array}
\]
Reflection at positive pairs uses monoidal functoriality
$\C\,\A_1 \todot \C\,\A_2 \todot \C\,(\A_1 \hattimes \A_2)$
called $\star$ by McBride and Paterson \cite[Section 7]{mcbridePaterson:applicative},
which for monads $\C$ can be defined by
$c_1 \star c_2 =
   \tjoin \; (\tmap\;(\lambda\,a_1.\
   \tmap\;(\lambda\,a_2.\ (a_1,a_2))
    \;c_2)
    \;c_1)$.
% $c_1 \star c_2 =
%    \tjoin \; (\tmap\;(\lambda\,a_1.\
%    \tjoin \; (\tmap\;(\lambda\,a_2.\ \treturn\,(a_1,a_2))
%     \;c_2))
%     \;c_1)$.
% $c \star d = \tjoin \; (\tmap\;(\lambda\,b.\ (c,\, \treturn\,b))\;d$.
% $\tzip\;c = \tjoin \comp \tmap\,(\lambda\,a.\ (c,\, \treturn\,a)$.

@@

Due to the eager splitting of positive hypotheses, reflection at type $P$
now lives in the graded monad $\Add P$ rather than $\Cov$.  Further,
as pattern matching may produce $n \geq 0$ cases, reflection cannot simply
produce a single positive semantic value; instead, one such value is needed
for every branch.  We implement
\fbox{$\reflect P : (\den P \hatto \J) \todot \Add P \J$}
as a higher-order function expecting a continuation $k$ which is
invoked for each generated branch with the semantic value of type $P$
constructed for this branch.
\[
\renewcommand{\left}{}
\renewcommand{\right}{}
\begin{array}{l@{~}lcl}
  \reflect {\patom} & k & = &
    \phyp \left( k\;\twk^{\patom}\,(\pvar\,\tzero) \right) \\
  \reflect {\Thunk N} & k & = &
    \nhyp \left( k\;\twk^N \left( \up N (\nvar\,\tzero) \right) \right)\\
  \reflect {\tyzero} & k & = & \nbranch \\
  \reflect {P_1 + P_2} & k & = & \bbranch
    \,\left( \treflect^{P_1}\,(\lambda\,\tau. \ k\,\tau \comp \iota_1) \right)
\\ &&& \qquad \quad \ \,
    \,\left( \treflect^{P_2}\,(\lambda\,\tau. \ k\,\tau \comp \iota_2) \right)
    \\
  \reflect{\tyone} & k & = & \nsplit\;(k\;\tid\;\ttempty) \\
  \reflect{P_1 \times P_2} & k & = & \bsplit
    \,\left(
      \treflect^{P_1} \left( \lambda\,\tau_1\,a_1. \
\\ &&& \qquad \ \
      \treflect^{P_2} \left( \lambda\,\tau_2\,a_2. \
\\ &&& \qquad \ \
        k\, (\tau_1 \rcomp \tau_2)\, (\tren\,\tau_2\,a_1, \ a_2)
        \right)
      \right)
    \right)
\end{array}
\]
Reflecting at a positive atomic type $\patom$ is the regular ending of a
reflection pass: we call continuation $k$ with a fresh variable
$\pvar\,\tzero$ of type $\patom$, making space for the variable
using $\twk^{\patom}$.
In case we end at type $\Thunk N$, we add a new variable
$\nvar\,\tzero$ of type $N$ and pass it to $k$, after full
$\eta$-expansion via $\up N$.  Two more endings are possible:
At type
$\tyzero$, we have reached an absurd case, meaning that no
continuation is necessary since we can conclude with
\emph{ex falso quodlibet}.
At type $\tyone$, there is no need to add a new variable, as values
of type $\tyone$ contain no information.  We simply pass the unit
value $\ttempty$ to $k$ in this case.
Reflecting at $P_1 + P_2$ generates two branches, which may result in
several uses of the continuation $k$.
In the first branch, we recursively reflect at $P_1$.  Its
continuation will receive a semantic value in $\den{P_1}$, which we
inject via $\iota_1$ into $\den{P_1 + P_2}$ to pass it to $k$.  The
second branch proceeds analogously.
Finally reflecting at $P_1 \times P_2$ means we first have to analyze
$P_1$, and in each of the generated branches we continue to analyze
$P_2$.  Thus $\reflect{P_2}$ is passed as a continuation to
$\reflect{P_1}$.  Each reflection phase gives us a semantic value
$a_i$ of type $P_i$, which we combine to a tuple before passing it to
$k$.  Note also that the context extension $\tau_1$ created in the
first phase needs to be composed with the context extension $\tau_2$
of the second phase to transport $k$ into the final context.  Further,
the value $a_1$ was constructed relative to the target of
$\tau_1$ and still needs
to be transported with $\tau_2$ before being paired up with $a_2$.

@@

The terms \fbox{$\Tm N \Gamma$} of the focused lambda calculus are the
ones of CBPV minus the positive eliminations ($\tsplit$, $\tcase$,
$\tabort$), the added negative variable rule ($\nvar$), and the
necessary changes to the binders $\tabs$ and $\tbind$.
\begin{gather*}
  \nru{\nvar}{N \in \Gamma}{\Tm N \Gamma}
\qquad
  \nru{\tabs}
      {\AddHypP P {\tTm\, N} \Gamma}
      {\TmP {P \To N} \Gamma}
\\[2ex]
  \nru{\tbind}
      {\TmP {\Comp P} \Gamma \qquad \AddHypP P {\tTm\,N} \Gamma}
      {\Tm N \Gamma}
\end{gather*}
Term interpretation
\fbox{$\denty\_ : \Tm N \Gamma \to \den \Gamma \todot \den N$}
shall be as for CBPV except that we need to exchange the
interpretation function for binders
%\fbox{$
\[
\fden\_ : \Tm N {(\cext
    \Gamma P)} \to \den \Gamma \todot \den{P \To N}
.\]
%$}.
Since a binder
for $P$ performs a maximal splitting on $P$ and takes the form of a
function defined by a case (and split) tree,
applying it to a value $v$ of type $P$ amounts to a complete
matching of $v$ against the case tree and binding the remaining
atomic and negative crumbs.  This matching can be defined for a
generic evaluation function of type
$\Evvv \J \Delta \Gamma = \den \Gamma \Delta \to \J\,\Delta$.
\[
\begin{array}{l@{~}l@{~}l@{~}lcl}
  \multicolumn 6 {l} {\tmatch ~~:~~ \den P \Delta \to
    \AddHypP P {\Evv \J \Delta} \todot \Evv \J \Delta }\\
  \tmatch & x & (\phyp\,e) & \gamma & = & e\,(\gamma,\; \treturn\,x) \\
  \tmatch & b & (\nhyp\,e) & \gamma & = & e\,(\gamma,\; b) \\
  \tmatch & a & \nbranch & \gamma & = & \tmagic\;a \\
  \tmatch & (\iota_1\;a_1) & (\bbranch\;e_1\;e_2) & \gamma & = &
    \tmatch\;a_1\;e_1\;\gamma \\
  \tmatch & (\iota_2\;a_2) & (\bbranch\;e_1\;e_2) & \gamma & = &
    \tmatch\;a_2\;e_2\;\gamma \\
  \tmatch & \ttempty & (\nsplit\,e) & \gamma & = & e\,\gamma \\
  \tmatch & (a_1,a_2) & (\bsplit\,e) & \gamma & = &
\\ \multicolumn 6 {@{\qquad}l}{
    \tmatch\;a_1\;(\tmap[\Add P]\,(\tmatch\;a_2)\;e)\;\gamma
}
\\
\end{array}
\]
With instantiations $\J = \den N$ and
$\denty\_ : \Tm N \todot \Evv {\den N} \Delta$,
the interpretation $\fden t$
of binder $t : \AddHypP P {\tTm\,N} \Gamma$
is defined as follows:
\begin{multline*}
  \fdent{t}
    {(\gamma : \den \Gamma \Delta)}
    \, (\tau : \Delta \sublist \Phi)
    \, (a : \den P \Phi)
    ~~=~~
\\ \qquad
    \tmatch
      \; a
      \; (\tmap[\Add P]\;\denty\_\;t)
      \; (\tren\,\tau\,\gamma)
\end{multline*}
This completes the definition of the normalization function
$\tnorm\, (t : \Tm N \Gamma)
  = \down N \dent t {\idenv\Gamma}$.
